---
title: "Week 3 Exercises"
author: "James Clark"
date: "March 2023"
output: pdf_document
header-includes:
- |
  ```{=latex}
  \usepackage{fvextra}
  \DefineVerbatimEnvironment{Highlighting}{Verbatim}{
    breaklines,
    commandchars=\\\{\}
  }
  \DefineVerbatimEnvironment{verbatim}{Verbatim}{
    breaklines,
    commandchars=\\\{\}
  }
  ```
---

```{r,include=FALSE}
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 60), tidy = TRUE)
```

Please complete all exercises below. You may use any library that we have covered in class UP TO THIS POINT.

```{r}
library(hash)
```

1)  Two Sum - Write a function named two_sum()

Given a vector of integers nums and an integer target, return indices of the two numbers such that they add up to target.

You may assume that each input would have exactly one solution, and you may not use the same element twice.

You can return the answer in any order.

Example 1:

Input: nums = [2,7,11,15], target = 9 Output: [0,1] Explanation: Because nums[0] + nums[1] == 9, we return [0, 1]. Example 2:

Input: nums = [3,2,4], target = 6 Output: [1,2] Example 3:

Input: nums = [3,3], target = 6 Output: [0,1]

Constraints:

2 \<= nums.length \<= 104 --109 \<= nums[i] \<= 109 --109 \<= target \<= 109 Only one valid answer exists.

*Note: For the first problem I want you to use a brute force approach (loop inside a loop)*

*The brute force approach is simple. Loop through each element x and find if there is another value that equals to target -- x*

*Use the function seq_along to iterate*

```{r}
two_sum <- function(nums_vector,target){
  #counter<-0 #initialize counter to count number of iterations
  
  if(is.numeric(nums_vector)){ #error handling to make sure vector is numeric, if one value in vector is not numeric, the vector will default to everything being that non-numeric type.
    for(i in seq_along(nums_vector)){
      second_value <- target-nums_vector[i]
      for(j in seq_along(nums_vector)){
        #counter<-counter+1 #add 1 to counter
        if(nums_vector[i]!=nums_vector[j] & nums_vector[j]==second_value){
          print(c(i,j))
         break    #assume that each input would have exactly one solution 
        }
      }
    }
    #cat("The number of iterations is ", counter,".\n",sep="")
  }else{
    stop("ERROR: One or more of your values in the vector is not numeric")
  }
  
}


# Test code
nums_vector <- c(5,7,12,34,6,10,8,9)
target <- 13
 
two_sum(nums_vector,target)
#expected answers
#[1] 1 7
#[1] 2 5
#[1] 5 2

```

```{r,include=FALSE}
two_sum_alt <- function(nums_vector,target){
  #Since addition is commutative nums_vector[i]+nums_vector[j]==nums_vector[j]+nums_vector[i], we can reduce calculations by just finding only c(i,j) solution instead of both c(i,j) and c(j,i)
  
  counter<-0 #initialize counter to count number of iterations
  if(is.numeric(nums_vector)){ #error handling to make sure vector is numeric
    for(i in seq_along(nums_vector[-c(length(nums_vector))])){ #sequence along vector except for last value, since you cant use same element twice
      second_value <- target-nums_vector[i]
      for(j in (i+1):length(nums_vector)){
        counter<-counter+1 #add 1 to counter
        if(nums_vector[j]==second_value){
          print(c(i,j))
         break    #assume that each input would have exactly one solution 
        }
      }
    }
    cat("The number of iterations is ", counter,".\n\n",sep="")
  }else{
    stop("ERROR: One or more of your values in the vector is not numeric")
  }
  
}
two_sum_alt(nums_vector,target)

two_sum_alt2 <- function(nums_vector,target){
  #alternate way to find solution using one loop and vectorization
  
  counter<-0 #initialize counter to count number of iterations
  if(is.numeric(nums_vector)){ #error handling to make sure vector is numeric
    for(i in seq_along(nums_vector)){
      counter<-counter+1 #add 1 to counter
      if(length(which(target-nums_vector[i]==nums_vector))>0){ #vectorized version of checking if there are any solutions
        print(c(i,which(target-nums_vector[i]==nums_vector)))  
      }
    }
    cat("The number of iterations is ", counter,".\n\n",sep="")
  }else{
    stop("ERROR: One or more of your values in the vector is not numeric")
  }
}

two_sum_alt2(nums_vector,target)

two_sum_alt3 <- function(nums_vector,target){
  #Second alternate way to find solution using one loop and vectorization and reducing calculations by finding only c(i,j) solution instead of both c(i,j) and c(j,i)
  
  temp_vector<-nums_vector #place holder vector to be used for while loop
  counter<-0 #initialize counter to count number of iterations
  if(is.numeric(nums_vector)){ #error handling to make sure vector is numeric
    while(length(temp_vector)>1){ #keep removing values checked and values that are part of solutions until only one is left. Since we need two values to sum, the last value in temp_vector does not need to be checked.
      counter<-counter+1 #add 1 to counter
      if(length(which(target-temp_vector[1]==nums_vector))>0){ #check if first entry in temp_vector is part of a solution
        print(c(which(nums_vector==temp_vector[1]),which(target-temp_vector[1]==nums_vector))) #if first entry of temp_vector is part of a solution, print ordered pair c(i,j)
        temp_vector<-temp_vector[-c(which(target-temp_vector[1]==temp_vector))]
        #after solution c(i,j) is found, this will remove checking for c(j,i) since we are assuming that each input would have exactly one solution
      }
      temp_vector<-temp_vector[-c(1)] #remove first entry of temp_vector after we finish checking if it is a solution
    }
    cat("The number of iterations is ", counter,".\n\n",sep="")
  }else{
    stop("ERROR: One or more of your values in the vector is not numeric")
  }
}

two_sum_alt3(nums_vector,target)

```

2)  Now write the same function using hash tables. Loop the array once to make a hash map of the value to its index. Then loop again to find if the value of target-current value is in the map.

*The keys of your hash table should be each of the numbers in the nums_vector minus the target.*

*A simple implementation uses two iterations. In the first iteration, we add each element's value as a key and its index as a value to the hash table. Then, in the second iteration, we check if each element's complement (target -- nums_vector[i]) exists in the hash table. If it does exist, we return current element's index and its complement's index. Beware that the complement must not be nums_vector[i] itself!*

```{r}

two_sum_hash <- function(nums_vector,target){
  if(is.numeric(nums_vector)){ #error handling to make sure vector is numeric
    h<-hash()
    #counter<-0 #initialize counter to count number of iterations
    
    for(i in seq_along(nums_vector)){
      #counter<-counter+1 #add 1 to counter
      h[nums_vector[i]]<-i
    } #build hash dictionary
    
    for(j in seq_along(nums_vector)){
      #counter<-counter+1 #add 1 to counter
      if(!is.null(h[[as.character(target-nums_vector[j])]])){ #checks to see if value we are looking for is in key of hash
        print(c(j,h[[as.character(target-nums_vector[j])]])) #print index j and index for value target-nums_vector[j], which are solutions
      }
    }
    #cat("The number of iterations is ", counter,".\n\n",sep="")
  }else{
    stop("ERROR: One or more of your values in the vector is not numeric")
  }
  
}


# Test code
nums_vector <- c(5,7,12,34,6,10,8,9)
target <- 15

#target<-13 #Used to compare to solutions in previous Problem #1
 
two_sum_hash(nums_vector,target)
#expected answers
#[1] 10  5
#[1] 8 7
#[1] 9 6
#[1]  5 10
#[1] 7 8
#[1] 6 9



```

```{r, include=FALSE}
two_sum_hash_2 <- function(nums_vector,target){
  #Since addition is commutative nums_vector[i]+nums_vector[j]==nums_vector[j]+nums_vector[i], we can reduce calculations by just finding c(i,j) solution instead of both c(i,j) and c(j,i)
  
  if(is.numeric(nums_vector)){ #error handling to make sure vector is numeric
    h<-hash()
    counter<-0 #initialize counter to count number of iterations
    
    for(i in seq_along(nums_vector)){
      counter<-counter+1 #add 1 to counter
      h[nums_vector[i]]<-i
    } #build hash dictionary
    
    temp_vector<-nums_vector #place holder vector to be used for while loop
    
    while(length(temp_vector)>1){#keep removing values checked and values that are part of solutions until only one is left. Since we need two values to sum, the last value in temp_vector does not need to be checked.
      counter<-counter+1 #add 1 to counter
      if(!is.null(h[[as.character(target-temp_vector[1])]])){#checks to see if value we are looking for is in key of hash
        print(c(which(temp_vector[1]==nums_vector),h[[as.character(target-temp_vector[1])]])) #print index j and index for value target-nums_vector[j], which are solutions
        temp_vector<-temp_vector[-c(which(target-temp_vector[1]==temp_vector))] #after solution c(i,j) is found, this will remove checking for c(j,i) since we are assuming that each input would have exactly one solution
      }
      temp_vector<-temp_vector[-c(1)] #remove first entry of temp_vector after we finish checking if it is a solution
    }
    cat("The number of iterations is ", counter,".\n\n",sep="")
  }else{
    stop("ERROR: One or more of your values in the vector is not numeric")
  }
}

two_sum_hash_2(nums_vector,target)
```
